---
title: "Longitudinal Mixed-Effects Modeling and Sensitivity Analyses in Simulated Parkinson's Disease Data"
author: "Dean Leonardi, Ph.D."
date: today
params:
  missing_rate: 0.25
editor: visual
---

## Abstract

A longitudinal analysis for Parkinson's patient UPDRS scores using MAR multiple imputation (MICE) sensitivity analysis and propensity score matching: a non-randomized, observational, simulated data study

## Executive Summary

This report demonstrates a reproducible longitudinal clinical reporting workflow using R and Quarto in a simulated Parkinson's disease observational study. Analyses include:

- longitudinal mixed-effects modeling,
- MAR multiple imputation sensitivity analysis (MICE),
- propensity score matching,
- treatment effect estimation,
- and automated HTML reporting.

This workflow is intended as a methodological demonstration and does not represent a regulatory submission analysis.

::: callout-note
## Key Features Demonstrated

- Longitudinal mixed-effects modeling
- Sensitivity analysis workflows
- Missing data handling
- Propensity score matching
- Automated statistical reporting
- Interactive HTML reporting with Quarto
- Reproducible research workflows
- Clinical-style statistical interpretation
:::

## Background

We assessed the efficacy of treatment in patients with Parkinson's disease (PD) using UPDRS (Unified Parkinson's Disease Rating Scale) scores. In PD research, the UPDRS is the gold standard for measuring motor and non-motor symptoms.

## Methods

This was a non-randomized, observational, 2-arm, simulated data study for adults with ages sampled from a normal distribution with mean 70 and standard deviation 12 years. Sex was sampled from a Bernoulli distribution with probability of being female 0.6. Treatment groups were assigned based on age and sex (non-random, simulating selection bias). We simulated longitudinal data for Parkinson's patients with UPDRS scores measured at baseline, 1 year, and 2 years. The two groups were control and treatment. Longitudinal analysis was used to track changes in UPDRS scores across multiple time points to determine if treatment slows or accelerates disease progression and identifies symptom trajectories (e.g., motor decline, non-motor worsening) and supports trial design and stratification. The study data was simulated so that older and more females tended to get treatment. The primary endpoint was the differences between treatment groups (treatment - control) in UPDRS scores at year 2. No safety analysis was conducted. The ITT analysis set included all patients in the simulated data set. Multiple imputation by chained equations (MICE) under a missing-at-random (MAR) assumption, with M = 25 completed datasets pooled via Rubin's Rules, and propensity score matching to balance age and sex were utilized for sensitivity analyses.

## Analysis

We simulate longitudinal data for Parkinson's patients with UPDRS scores measured at baseline, 1 year, and 2 years. Two groups: Control and Treatment.

```{r}
#| label: load-packages
#| include: false


library(tidyverse)
library(dplyr)
library(gtsummary)
library(gt)
library(lme4)
library(lmerTest)
library(emmeans)
library(tableone)  # Baseline comparison tables (SMD) or Balance checking
library(MatchIt)   # Performs propensity score matching
library(cobalt)    # Performs propensity score matching
library(mice)        # multiple imputation
library(broom.mixed) # required for mice::pool() to work with lmer models
library(naniar)
```

```{r}
#| label: simulate-data
set.seed(456)
# Non-randomized observational study: older and more females tend to get treatment
n_patients <- 400
patients <- 1:n_patients
times <- c(0, 1, 2)  # baseline, 1 year, 2 years

# Non-random group assignment: function for treatment probability
prob_treatment <- function(age, sex) {
  plogis(-1 + 0.05*(age - 65) + 0.6*(sex == "F"))
}

# Create baseline covariates first
baseline_covariates <- data.frame(
  Patient = as.factor(patients),
  Age = rnorm(n_patients, mean = 70, sd = 12),  # Older population
  Sex = sample(c("M", "F"), n_patients, replace = TRUE, prob = c(0.4, 0.6))  # More females
)

# Assign groups based on age and sex (non-random, simulating selection bias)
baseline_covariates$Group <- as.factor(sapply(1:n_patients, function(i) {
  if (runif(1) < prob_treatment(baseline_covariates$Age[i], baseline_covariates$Sex[i])) {
    "Treatment"
  } else {
    "Control"
  }
}))

# Create longitudinal data frame
data <- expand.grid(Patient = patients, Time = times)
data$Patient <- as.factor(data$Patient)
data$Time <- as.factor(data$Time)

# Merge with baseline covariates
data <- data %>% left_join(baseline_covariates, by = "Patient")

# Simulate UPDRS scores
data$UPDRS <- rnorm(nrow(data), mean = 35 + 5 * as.numeric(as.character(data$Time)) +
                      ifelse(data$Group == "Treatment", -1.5, 0), sd = 6)

# Parameterization of missing data
data$UPDRS <- ifelse(
  runif(nrow(data)) < params$missing_rate &
  data$Time != "0",
  NA,
  data$UPDRS
)

data |>
  head() |>
  gt() |>
  tab_header(title = "Simulated Data — First Six Rows") |>
  tab_options(table.font.size = px(12))

# Add at the bottom of the simulate-data chunk
data_wide <- data |>
  select(Patient, Time, Age, Sex, Group, UPDRS) |>
  pivot_wider(
    id_cols      = c(Patient, Age, Sex, Group),
    names_from   = Time,
    values_from  = UPDRS,
    names_prefix = "UPDRS_t"
  )

```

### Computational Reproducibility

Random seeds were pre-specified for all simulation and matching procedures to ensure reproducible analytical outputs.

### Demographics

```{r}
#| label: demo-table
#| echo: false

data |> filter(Time == "0") |>
  tbl_summary(
    by = Group,
    include = c(Age, Sex, UPDRS),
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{n} / {N} ({p}%)"
    ),
    digits = all_continuous() ~ 2,
    missing_text = "(Missing)"
  ) |>
  add_overall() |>
  as_gt() |>
  tab_header(
    title = "Table 1. Summary of Demographics and Baseline Characteristics"
  ) |>
  gt::tab_source_note(gt::md("*Data is simulated*"))
```

### Primary Analysis

Fit a linear mixed effects model to estimate least squares means differences between treatment groups at year 2.

```{r}
#| label: lme-model

data$Group <- factor(data$Group,
                     levels = c("Control", "Treatment"))
m1 <- lmer(UPDRS ~ Group*Time + Age + (1|Patient), data=data)

# ── Convergence check ─────────────────────────────────────────────────────────
# Surface any optimizer warnings in the rendered output so they are not silently
# swallowed. A NULL result here means the model converged without issues.
conv_msgs <- m1@optinfo$conv$lme4$messages
if (!is.null(conv_msgs)) {
  warning("Convergence warning(s): ", paste(conv_msgs, collapse = "; "))
} else {
  message("Model converged without warnings.")
}

source("R/utils.R")


u4 <- extract_emm_table(m1)
u4 |>
  gt() |>
  format_gt_table() |>
  tab_header(
    title = "Table 2. Summary of Linear Mixed-Effects Model Predicting UPDRS by Treatment and Age at Year 2"
  ) |>
  tab_footnote(
    footnote = "CI: 95% Confidence Interval; Formula: UPDRS ~ Group * Time + Age + (1 | Patient)"
  )

# emmeans estimates
est_yr2 <- u4 |> filter(grepl("Treatment - Control", Group)) |> pull(Estimate)
ci_yr2  <- u4 |> filter(grepl("Treatment - Control", Group)) |> pull(CI)

```

```{r}
#| label: icc
#| echo: true

# ── Intraclass Correlation Coefficient (ICC) ──────────────────────────────────
# ICC = σ²_u / (σ²_u + σ²): proportion of total variance attributable to
# between-patient differences. A high ICC (> 0.5) confirms that patients differ
# substantially at baseline and validates the need for a mixed model — standard
# regression would underestimate standard errors by treating within-person
# observations as independent.

vc       <- as.data.frame(VarCorr(m1))
var_u    <- vc$vcov[vc$grp == "Patient"]   # between-patient variance
var_eps  <- vc$vcov[vc$grp == "Residual"]  # within-patient (residual) variance
icc_val  <- var_u / (var_u + var_eps)

tibble(
  Component              = c("Between-patient (σ²_u)", "Within-patient / Residual (σ²)", "ICC"),
  Variance               = c(round(var_u, 3), round(var_eps, 3), NA_real_),
  Value                  = c(NA_character_, NA_character_, round(icc_val, 3))
) |>
  gt() |>
  tab_header(
    title    = "Table 2b. Variance Components and Intraclass Correlation Coefficient",
    subtitle = "Random-intercept LME — UPDRS ~ Group * Time + Age + (1 | Patient)"
  ) |>
  cols_label(
    Component = "Variance Component",
    Variance  = "Estimated Variance",
    Value     = "ICC"
  ) |>
  tab_footnote(
    footnote = paste0(
      "ICC = σ²_u / (σ²_u + σ²). ",
      "ICC ≈ 0.1: weak within-patient correlation. ",
      "ICC ≈ 0.3: moderate. ",
      "ICC > 0.5: strong longitudinal dependence — standard regression would substantially underestimate SEs."
    )
  ) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_options(
    table.font.size           = px(13),
    heading.title.font.size   = px(14),
    heading.title.font.weight = "bold",
    table.width               = pct(100)
  )
```

The longitudinal mixed-effects model demonstrated lower estimated UPDRS scores in the treatment group at Year 2 relative to control, consistent with simulated treatment benefit.

```{r}
#| label: longitudinal-profile-plot
#| fig-cap: "Figure 1. Longitudinal mean UPDRS profiles by treatment group."
#| fig-width: 8
#| fig-height: 5
#| warning: false
#| message: false

# Summary statistics by treatment group and visit
plot_data <- data %>%
  group_by(Group, Time) %>%
  summarize(
    Mean = mean(UPDRS, na.rm = TRUE),
    SD = sd(UPDRS, na.rm = TRUE),
    N = sum(!is.na(UPDRS)),
    SE = SD / sqrt(N),
    Lower = Mean - 1.96 * SE,
    Upper = Mean + 1.96 * SE,
    .groups = "drop"
  )

# Convert Time to numeric for plotting
plot_data$Time <- as.numeric(as.character(plot_data$Time))

# Longitudinal profile plot
ggplot(plot_data,
       aes(x = Time,
           y = Mean,
           color = Group,
           group = Group)) +

  geom_line(linewidth = 1.2) +

  geom_point(size = 3) +

  geom_errorbar(
    aes(ymin = Lower, ymax = Upper),
    width = 0.08,
    linewidth = 0.7
  ) +

  scale_x_continuous(
    breaks = c(0, 1, 2),
    labels = c("Baseline", "Year 1", "Year 2")
  ) +

  labs(
    title = "Longitudinal Mean UPDRS Profiles by Treatment Group",
    subtitle = "Observed Mean Scores with 95% Confidence Intervals",
    x = "Visit",
    y = "Mean UPDRS Score",
    color = "Treatment Group"
  ) +

  theme_minimal(base_size = 13) +

  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "top"
  )
```

```{r}
#| label: residual-diagnostics
#| fig-cap: "Figure 2. Residual diagnostics for the primary mixed-effects model."
#| fig-width: 10
#| fig-height: 8
#| warning: false
#| message: false

# ── Four-panel diagnostic layout ─────────────────────────────────────────────
# (1) Residuals vs Fitted   — checks linearity and homoscedasticity
# (2) Q-Q plot of residuals — checks normality of marginal residuals
# (3) Q-Q plot of BLUPs     — checks normality of random-intercept estimates
# (4) Scale-Location        — checks homoscedasticity (sqrt |residuals| vs fitted)

par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))

# Panel 1: Residuals vs Fitted
plot(
  fitted(m1), resid(m1),
  xlab = "Fitted Values", ylab = "Residuals",
  main = "Residuals vs Fitted",
  pch  = 16, cex = 0.6, col = "steelblue"
)
abline(h = 0, lty = 2, col = "grey40")
lines(lowess(fitted(m1), resid(m1)), col = "tomato", lwd = 1.5)

# Panel 2: Q-Q plot of marginal residuals
qqnorm(
  resid(m1),
  main = "Q-Q Plot: Marginal Residuals",
  pch  = 16, cex = 0.6, col = "steelblue"
)
qqline(resid(m1), col = "tomato", lwd = 1.5)

# Panel 3: Q-Q plot of random effects (BLUPs)
blups <- ranef(m1)$Patient[, 1]
qqnorm(
  blups,
  main = "Q-Q Plot: Random Intercepts (BLUPs)",
  pch  = 16, cex = 0.7, col = "steelblue"
)
qqline(blups, col = "tomato", lwd = 1.5)

# Panel 4: Scale-Location
plot(
  fitted(m1), sqrt(abs(resid(m1))),
  xlab = "Fitted Values", ylab = expression(sqrt("|Residuals|")),
  main = "Scale-Location",
  pch  = 16, cex = 0.6, col = "steelblue"
)
lines(lowess(fitted(m1), sqrt(abs(resid(m1)))), col = "tomato", lwd = 1.5)

par(mfrow = c(1, 1))
```

Residual diagnostics show no systematic violations of model assumptions. The Q-Q plots assess normality of marginal residuals (Panel 2) and random-intercept BLUPs (Panel 3); departures from the reference line would suggest distributional misspecification. The residuals vs. fitted and scale-location plots (Panels 1 and 4) assess linearity and homoscedasticity; the loess smoother should track near zero with no fan-shaped pattern.

## Findings

Four hundred patients were assigned to receive control (n=`r sum(baseline_covariates$Group=="Control")`), or Treatment (n=`r sum(baseline_covariates$Group=="Treatment")`). Treatment reduced UPDRS score versus control (treatment 43.35 vs control 46.27; difference `r round(est_yr2, 2)`, 95% CI `r ci_yr2`).

## Interpretation

In patients with Parkinson's disease, treatment reduced the UPDRS score at Year 2. These findings are based on simulated data and are intended as a methodological demonstration only.

### Missing Data Analysis

Under a Missing At Random (MAR) assumption, missing UPDRS outcomes are predicted by observed covariates (Age, Sex, Group) and observed UPDRS values at other timepoints. Multiple imputation by chained equations (MICE) with predictive mean matching was used to generate M = 25 completed datasets. Fixed-effect estimates from each dataset were combined using Rubin's Rules to produce a single pooled estimate and standard error that properly accounts for within- and between-imputation uncertainty.

#### Table 3a. Missingness Summary Prior to Imputation

```{r}
#| label: missingness-summary
#| echo: false
#| message: false
#| warning: false

# Missingness summary table
data_wide |>
  summarise(
    `N patients`           = n(),
    `Missing — Year 1 (n)` = sum(is.na(UPDRS_t1)),
    `Missing — Year 2 (n)` = sum(is.na(UPDRS_t2)),
    `% Missing — Year 1`   = paste0(round(100 * mean(is.na(UPDRS_t1)), 1), "%"),
    `% Missing — Year 2`   = paste0(round(100 * mean(is.na(UPDRS_t2)), 1), "%")
  ) |>
  gt() |>
  tab_header(
    title    = "Table 3a. Missingness Summary Prior to Imputation",
    subtitle = "UPDRS_t0 (baseline) is fully observed by design"
  ) |>
  tab_options(
    table.font.size           = px(13),
    heading.title.font.size   = px(14),
    heading.title.font.weight = "bold",
    table.width               = pct(100)
  )
```

#### Figure 3a

```{r}
#| label: missingness-pattern
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Figure 3a. Missing data pattern across patients and visits."
#| fig-width: 10
#| fig-height: 5
#| out-width: "90%"
#| fig-align: center

naniar::vis_miss(
  data_wide |> dplyr::select(-Patient),
  warn_large_data = FALSE
) +
  theme(
    axis.text.x.top = element_text(angle = -45, hjust = 0, vjust = 0, size = 9),
    plot.margin     = margin(t = 20, r = 10, b = 10, l = 10, unit = "pt")
  )
```

```{r}
#| label: mice-imputation
#| message: false
#| warning: false
#| cache: true
#| cache.extra: !expr list(nrow(data_wide), ncol(data_wide))

# ── 1. Prepare data for mice ──────────────────────────────────────────────────
# Remove Patient ID before passing to mice (high-cardinality ID should not
# be a predictor). Store IDs separately and re-attach after completion.

patient_ids <- data_wide |> pull(Patient)

data_mice <- data_wide |>
  select(-Patient) |>
  mutate(
    Sex   = factor(Sex,   levels = c("M", "F")),
    Group = factor(Group, levels = c("Control", "Treatment"))
  )

# ── 2. Configure predictor matrix and imputation methods ──────────────────────
ini  <- mice(data_mice, maxit = 0, print = FALSE)
meth <- ini$method
pred <- ini$predictorMatrix

# Baseline UPDRS is fully observed — exclude from imputation target
# but retain as a predictor for Year 1 and Year 2 imputation
meth["UPDRS_t0"] <- ""
pred["UPDRS_t0", ] <- 0

# pmm (predictive mean matching) is appropriate for bounded continuous scores:
# it draws observed values close to the linear prediction, preserving the
# empirical distribution and avoiding impossible out-of-range imputations
meth[c("UPDRS_t1", "UPDRS_t2")] <- "pmm"

# ── 3. Run mice ────────────────────────────────────────────────────────────────
set.seed(2026)
imp_mar <- mice(
  data_mice,
  m               = 25,    # 25 datasets: adequate for ~25% missingness
  method          = meth,
  predictorMatrix = pred,
  maxit           = 15,    # iterations per dataset until convergence
  print           = FALSE
)
```

```{r}
#| label: mice-convergence
#| fig-cap: "Figure 3b. Convergence diagnostics for mice imputation. Mean and SD of imputed values should mix well (no trend) across iterations."
#| fig-width: 8
#| fig-height: 4
#| message: false
#| warning: false

# Trace plots: flat, well-mixed chains indicate convergence
plot(imp_mar, c("UPDRS_t1", "UPDRS_t2"), layout = c(2, 2))
```

```{r}
#| label: mice-fit-pool
#| message: false
#| warning: false

# ── 4. Fit LMER on each completed dataset ─────────────────────────────────────
# For each of the M imputed wide datasets:
#   (a) re-attach Patient IDs
#   (b) reshape wide → long
#   (c) fit the same LMER specification as the primary analysis

fit_on_completed <- function(i) {

  wide_i <- mice::complete(imp_mar, action = i) |>
    mutate(Patient = patient_ids)                  # restore patient ID

  long_i <- wide_i |>
    pivot_longer(
      cols         = starts_with("UPDRS_t"),
      names_to     = "Time",
      names_prefix = "UPDRS_t",
      values_to    = "UPDRS"
    ) |>
    mutate(
      Time  = factor(Time,  levels = c("0", "1", "2")),
      Group = factor(Group, levels = c("Control", "Treatment"))
    )

  lmer(UPDRS ~ Group * Time + Age + (1 | Patient), data = long_i)
}

models_mar <- lapply(seq_len(imp_mar$m), fit_on_completed)

# ── 5. Pool fixed effects via Rubin's Rules ───────────────────────────────────
# mice::pool() uses broom.mixed::tidy() to extract estimates from each lmerMod
pooled_mar  <- mice::as.mira(models_mar) |> mice::pool()
pooled_summ <- summary(pooled_mar, conf.int = TRUE)

# ── 6. Display pooled fixed effects table ─────────────────────────────────────
pooled_summ |>
  as_tibble() |>
  rename(
    Parameter = term,
    Estimate  = estimate,
    SE        = std.error,
    df        = df,
    p         = p.value,
    Lower     = `2.5 %`,
    Upper     = `97.5 %`
  ) |>
  mutate(
    across(c(Estimate, SE, Lower, Upper), \(x) round(x, 3)),
    CI = paste0("(", round(Lower, 2), ", ", round(Upper, 2), ")"),
    p  = case_when(
           p < 0.001 ~ "< 0.001",
           TRUE      ~ as.character(round(p, 3))
         )
  ) |>
  select(Parameter, Estimate, SE, CI, df, p) |>
  gt() |>
  tab_header(
    title    = "Table 3b. Pooled Fixed Effects — MAR Sensitivity Analysis",
    subtitle = "MICE with Predictive Mean Matching; m = 25; Rubin's Rules"
  ) |>
  cols_label(
    Parameter = "Parameter",
    Estimate  = "Estimate",
    SE        = "Std. Error",
    CI        = "95% CI",
    df        = "Pooled df",
    p         = "p-value"
  ) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_footnote(
    footnote = "Pooled using Rubin's Rules. Formula: UPDRS ~ Group × Time + Age + (1 | Patient). FMI reported in Table 3c."
  ) |>
  tab_options(
    table.font.size           = px(13),
    heading.title.font.size   = px(14),
    heading.title.font.weight = "bold",
    table.width               = pct(100)
  )
```

```{r}
#| label: mice-emmeans-pool
#| message: false
#| warning: false

# ── 7. Pooled treatment effect at Year 2 formatted like primary analysis ──────
# emmeans is computed for each imputed model, then estimates and SEs are
# combined using Rubin's Rules. The Satterthwaite df from emmeans is used
# in the Barnard-Rubin (1999) degrees-of-freedom approximation.

pool_emmeans_contrast <- function(models, time_level = "2") {

  # Extract Treatment - Control contrast at the requested timepoint
  extract_one <- function(mod) {
    em  <- emmeans(mod, ~ Group | Time)
    ct  <- contrast(em, method = "trt.vs.ctrl", ref = "Control")
    row <- as.data.frame(ct) |> filter(Time == time_level)
    list(
      est = row$estimate,
      se  = row$SE,
      df  = row$df      # Satterthwaite df
    )
  }

  res   <- lapply(models, extract_one)
  M     <- length(res)
  ests  <- sapply(res, `[[`, "est")
  ses   <- sapply(res, `[[`, "se")
  dfs   <- sapply(res, `[[`, "df")

  # Rubin's Rules
  Q_bar <- mean(ests)                          # pooled estimate
  U_bar <- mean(ses^2)                         # average within-imputation variance
  B     <- var(ests)                           # between-imputation variance
  T_var <- U_bar + (1 + 1 / M) * B            # total variance
  SE_p  <- sqrt(T_var)

  # Fraction of missing information (FMI)
  r   <- (1 + 1 / M) * B / U_bar
  fmi <- r / (1 + r)

  # Barnard-Rubin (1999) pooled degrees of freedom
  df_old <- mean(dfs)
  df_m   <- (M - 1) / r^2
  df_obs <- (df_old + 1) / (df_old + 3) * df_old * (1 - r)
  df_p   <- 1 / (1 / df_m + 1 / df_obs)

  ci <- Q_bar + qt(c(0.025, 0.975), df = df_p) * SE_p

  tibble(
    Contrast    = "Treatment − Control",
    Time        = paste0("Year ", time_level),
    Estimate    = round(Q_bar, 3),
    SE          = round(SE_p,  3),
    CI          = paste0("(", round(ci[1], 2), ", ", round(ci[2], 2), ")"),
    FMI         = round(fmi, 3)
  )
}

contrast_pooled <- pool_emmeans_contrast(models_mar, time_level = "2")

contrast_pooled |>
  gt() |>
  tab_header(
    title    = "Table 3c. Pooled Treatment Effect at Year 2 — MAR Sensitivity Analysis",
    subtitle = "MICE / Predictive Mean Matching; m = 25; Rubin's Rules"
  ) |>
  cols_label(
    Contrast = "Contrast",
    Time     = "Timepoint",
    Estimate = "Estimate",
    SE       = "Std. Error",
    CI       = "95% CI",
    FMI      = "FMI"
  ) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_footnote(
    footnote = paste0(
      "FMI: Fraction of Missing Information — proportion of total variance ",
      "attributable to missing data. Higher values indicate stronger sensitivity ",
      "to the MAR assumption. df computed via Barnard-Rubin (1999) approximation."
    )
  ) |>
  tab_options(
    table.font.size           = px(13),
    heading.title.font.size   = px(14),
    heading.title.font.weight = "bold",
    table.width               = pct(100)
  )
```

Results from the MAR multiple imputation sensitivity analysis are presented in Tables 3b and 3c. Consistency with the primary analysis estimates would support robustness of the primary findings to the missing-at-random assumption. The Fraction of Missing Information (FMI) quantifies the proportion of uncertainty attributable to missing data; values below 0.10 are generally considered low.

### Propensity Score Matching

Assuming the groups are not randomized, use propensity score matching to balance covariates of age and sex. Based on the matched data, fit the longitudinal mixed-effects model and estimate least squares means differences between treatment groups at year 2.

```{r}
#| label: lme-psm
set.seed(2026)

# Propensity score matching and evaluation
vars <- c("Age", "Sex")
table_before <- CreateTableOne(vars = vars,
                               strata = "Group",
                               data = baseline_covariates,
                               test = FALSE)

#print(table_before, smd = TRUE)

print(table_before, smd = TRUE, printToggle = FALSE) |>
  as.data.frame() |>
  tibble::rownames_to_column("Characteristic") |>
  gt() |>
  format_gt_table() |>           # your existing utility
  tab_header(title = "Table 4. Balance Before Matching") |>
  tab_footnote("SMD < 0.10 indicates acceptable balance.")



match_model2 <- matchit(Group ~ Age + Sex,
                        data = baseline_covariates,
                        method = "nearest",
                        link="logit",
                        distance="glm")

matched_data2 <- match.data(match_model2)
#table(matched_data2$Group)


table_after <- CreateTableOne(vars = vars,
                              strata = "Group",
                              data = matched_data2,
                              test = FALSE)

#print(table_after, smd = TRUE)

print(table_after, smd = TRUE, printToggle = FALSE) |>
  as.data.frame() |>
  tibble::rownames_to_column("Characteristic") |>
  gt() |>
  format_gt_table() |>           # your existing utility
  tab_header(title = "Table 5. Balance After Matching") |>
  tab_footnote("SMD < 0.10 indicates acceptable balance.")

# Subset orginal data based on patients identified by Propensity score matching
matched_pid <- matched_data2 |> select(Patient)
matched_data_final <- left_join(matched_pid, data, by = "Patient")


# Generate linear mixed effects model on matched population
matched_data_final$Group <- factor(
    matched_data_final$Group,
    levels = c("Control", "Treatment"))

m3 <- lmer(UPDRS ~ Group*Time + Age + (1|Patient), data=matched_data_final)

w4 <- extract_emm_table(m3)
w4 |>
  gt() |>
  format_gt_table() |>
  tab_header(
    title = "Table 6. Summary of Linear Mixed-Effects Model Predicting UPDRS by Treatment and Age at Year 2",
    subtitle = "Propensity Score Matching"
  ) |>
  tab_footnote(
    footnote = "CI: 95% Confidence Interval; Formula: UPDRS ~ Group * Time + Age + (1 | Patient)"
  )
```

Propensity score matching improved baseline covariate balance between treatment groups while preserving estimated treatment differences at Year 2.

```{r}
#| label: covbal-plot
#| fig-cap: "Figure 4. Covariate balance before and after propensity score matching."
#| fig-width: 8
#| fig-height: 5
#| warning: false
#| message: false
# Love Plot
love.plot(
  match_model2,
  binary = "std",
  thresholds = c(m = 0.1),
  abs = TRUE,
  var.order = "unadjusted",
  stars = "raw"
)
```

## Conclusions

This report demonstrates a reproducible and automated longitudinal clinical reporting workflow using modern R and Quarto methodologies. The analysis integrates mixed-effects modeling, missing data sensitivity analyses, and propensity score matching within a publication-style HTML reporting framework suitable for exploratory clinical and real-world evidence applications.

The workflow illustrates approaches commonly used in pharmaceutical and biotechnology clinical development environments for statistical analysis, reproducibility, and reporting automation.

## Limitations

This report uses simulated observational data for methodological demonstration purposes. Results should not be interpreted clinically. Missing data sensitivity analyses use MICE with predictive mean matching under a missing-at-random (MAR) assumption. Whether the MAR assumption holds cannot be verified from the observed data alone; MNAR sensitivity analyses (e.g., reference-based methods using the rbmi package) would be required for regulatory-grade submissions.

## Reproducibility

This report was generated using a reproducible Quarto workflow in R.\
All analyses, figures, tables, and statistical outputs were generated programmatically from simulated source data.

Key components include:

- automated data simulation,
- longitudinal mixed-effects modeling,
- MAR multiple imputation using MICE (m = 25, Rubin's Rules),
- propensity score matching,
- reproducible statistical reporting,
- and parameterized report generation.

The workflow is designed to reflect modern clinical reporting and real-world evidence analysis practices used in pharmaceutical and biotechnology settings.

## Appendix

```{r}
#| label: appendix
sessionInfo()
```
