---
title: "Time-to-Event Endpoints — Simulated Oncology RWE"
subtitle: "Risk-Set Adjusted KM, Cox PH, IPW/AIPW, and E-value Sensitivity"
author: "Dean Leonardi"
date: today
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(survival)
library(survminer)
library(simsurv)
library(ggsurvfit)
library(gt)
library(gtsummary)
library(WeightIt)
library(sandwich)
library(lmtest)
library(EValue)

source("R/simulate_oncology.R")
```

This vignette demonstrates a reproducible RWE survival analysis pipeline applied to a simulated oncology RCT (n = 400) with overall survival (OS) as the primary endpoint. Methods progress from descriptive Kaplan–Meier estimation through multivariable Cox regression, inverse probability weighting (IPW) for the average treatment effect (ATE), and E-value sensitivity analysis for unmeasured confounding — mirroring regulatory RWE practice for oncology submissions.

## Risk-Set Adjusted Kaplan–Meier

```{r}
#| label: km-plot
#| fig-cap: "Figure 1. Kaplan–Meier Overall Survival by Treatment Arm."
#| fig-width: 8
#| fig-height: 6
#| warning: false
#| message: false

fit <- survfit(Surv(os_time, os_event) ~ trt, data = data_onc)
ggsurvfit(fit) +
  add_confidence_interval() +
  add_risktable() +
  add_pvalue(caption = "Log-rank {p.value}") +
  scale_color_manual(values = c("0" = "#E34234", "1" = "#2E5F8A"),
                     labels = c("Control", "Treatment")) +
  labs(title = "Overall Survival by Treatment Arm",
       x = "Time (months)", y = "Survival Probability")
```

## Cox Proportional Hazards Model

```{r}
#| label: cox-model
#| tbl-cap: "Table 1. Multivariable Cox Proportional Hazards Model."
#| warning: false
#| message: false

cox_os <- coxph(
  Surv(os_time, os_event) ~ trt + age + ecog + stage,
  data = data_onc, ties = "efron"
)

broom::tidy(cox_os, exponentiate = TRUE, conf.int = TRUE) |>
  gt() |>
  fmt_number(columns = c(estimate, conf.low, conf.high), decimals = 2) |>
  tab_header(title = "Cox Proportional Hazards Model") |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_footnote(footnote = "Multivariable Cox model (trt + age + stage + ecog).") |>
  tab_options(
    table.font.size           = px(13),
    heading.title.font.size   = px(14),
    heading.title.font.weight = "bold",
    table.width               = pct(100)
  )
```

```{r}
#| label: schoenfeld-residuals
#| fig-cap: "Figure 2. Schoenfeld residuals test for PH assumption."
#| fig-width: 8
#| fig-height: 5
#| warning: false
#| message: false

ph_test <- cox.zph(cox_os)
ggcoxzph(ph_test)
```

```{r}
#| label: forest-plot
#| fig-cap: "Figure 3. Forest Plot — HRs with 95% CIs for all covariates."
#| fig-width: 8
#| fig-height: 5
#| warning: false
#| message: false

ggforest(cox_os, data = data_onc,
         main     = "Forest Plot — OS",
         fontsize = 0.85)
```

The adjusted HR for treatment = `r round(exp(coef(cox_os)["trt"]), 2)` (95% CI: `r round(exp(confint(cox_os)["trt", 1]), 2)`, `r round(exp(confint(cox_os)["trt", 2]), 2)`). A HR \< 1 favors treatment (reduced hazard of death); a HR \> 1 favors control. The Schoenfeld residuals plot above assesses the proportional hazards assumption — non-random trends over time would indicate a PH violation requiring model extension (e.g., time-varying coefficients or stratification).

## IPW / AIPW for ATE with Robust SE

Doubly robust estimation on oncology data.

- WeightIt: propensity score model (trt \~ age + stage + ecog), estimand="ATE"
- IPW-weighted Cox model for ATE with sandwich SE (regulatory practice)
- E-value for the primary IPW HR (unmeasured confounding sensitivity)

```{r}
#| label: ipw-aipw
#| tbl-cap: "Table 2. IPW-Weighted Cox Model with E-value Sensitivity Analysis."
#| warning: false
#| message: false

# IPTW model
w_out   <- weightit(trt ~ age + stage + ecog, data = data_onc,
                    method = "glm", estimand = "ATE")
cox_ipw <- coxph(Surv(os_time, os_event) ~ trt,
                 data = data_onc, weights = w_out$weights, robust = TRUE)

# Robust HR and CI
s    <- summary(cox_ipw)
beta <- coef(cox_ipw)["trt"]
se   <- s$coefficients["trt", "robust se"]
hr   <- exp(beta)
lo   <- exp(beta - qnorm(0.975) * se)
hi   <- exp(beta + qnorm(0.975) * se)

# E-values — evalues.HR() returns a matrix; extract numerically before building tibble
ev_mat <- evalues.HR(est = hr, lo = lo, hi = hi, rare = FALSE)
e_est  <- as.numeric(ev_mat[1, 1])
e_ci   <- as.numeric(ev_mat[1, 2])

results_tbl <- tibble(
  Covariate       = "trt",
  HR              = hr,
  Lower95CI       = lo,
  Upper95CI       = hi,
  EValue_Estimate = e_est,
  EValue_CI_Limit = e_ci
)

results_tbl |>
  gt() |>
  fmt_number(
    columns  = c(HR, Lower95CI, Upper95CI, EValue_Estimate, EValue_CI_Limit),
    decimals = 2
  ) |>
  cols_label(
    Covariate       = "Treatment",
    HR              = "Hazard Ratio",
    Lower95CI       = "95% CI Lower",
    Upper95CI       = "95% CI Upper",
    EValue_Estimate = "E-value",
    EValue_CI_Limit = "E-value (CI Limit)"
  ) |>
  tab_header(title = "IPW-Weighted Cox Model with E-value Sensitivity Analysis") |>
  tab_footnote(footnote = "ATE IPTW model using age, stage, and ECOG score.")
```

IPW-weighted Cox model for ATE with sandwich SE resulted in a HR = `r round(hr, 2)` (95% CI: `r round(lo, 2)`, `r round(hi, 2)`). The E-value quantifies the minimum strength of association an unmeasured confounder would need with both treatment and outcome to explain away the observed effect — larger values indicate greater robustness to unmeasured confounding.
